---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r echo=F, include=F}
library(ggplot2)
library(ggforce)
theme_set(theme_bw())
```


# Case study: Mediterranean fruit fly (*C. Capitata*)

TODO: Probabilistic graphical model of the entire model (with a box to demarcate the "internal" components from the components on which a prior distribution is set).

## Introduction

TODO: Rewrite this intro in light of the model proposal chapter

In the previous chapter, I proposed to address a gap in the literature by providing an elaborate model for inferring the eradication of an incipient invasive population of Tephritid fruit flies. In this chapter, I present an illustrative model of medfly surveillance after an hypothetical invasion. I use a simplified model of Medfly population dynamics. However, for various species of Tephritid fruit fly (medfly included) detailed models exist. A benefit of the proposed method is that it can easily incorporate almost any model of medfly dynamics.^[Note, though, that the sampling method I use may not be appropriate in all cases. When the model predicts that the population size "explodes", then the rejection rate for the sampling algorithm may become very high, causing the algorithm to be highly inefficient.]

TODO:

* Note that this analysis is primarily illustrative. Performing a more specific analysis requires access to confidential data.
* Note that the method can be used for a real scenario - we only need to change the priors, data and locations of the traps.
* Note that I use a simplified model of Medfly dynamics for illustrative purposes - but the method can easily incorporate more complex ABS models, and cite those models.

TODO: Write one paragraph outlining this chapter

## Medfly (Ceratitis capitata)

### Medfly are economically important

Mediterranean fruit fly (*Ceratitis Capitata*) or *medfly* are a particularly salient species of tephritid fruit fly. Medflies have high invasive potential, as it can adapt to a relatively large range of climates and environments, and is known to have the capability to infest the fruits of over 300 species of plants (Ibid.). Recently, an incursion of medfly in Adelaide, South Australia, prompted a large scale eradication effort. This comprised in part of hiring 350 special-purpose staff that set over 13,000 additional traps, and collected over 350 tonnes of fruit. The scale of the response to this outbreak indicates the perceived economic significance of this fruit fly species. 

### Medfly are cryptic

Medfly are very hard to detect at low levels. Monitoring for medfly is typically performed with the aid of lured traps (namely so-called Lynfield or Jackson traps). These traps are relatively ineffective for detecting medfly. For example, one study from the Adelaide metro area trapping grid found that only 0.02% of flies were recaptured from a release of 38.8 million flies. Further, medfly are known to have low dispersals across space. This means that low-lying populatons of flies may go undetected across generations. <https://onlinelibrary-wiley-com.virtual.anu.edu.au/doi/pdfdirect/10.1111/j.1570-7458.2006.00415.x>

## Data

In a Bayesian model, various sources of data can be used either to perform inference on parameters (i.e. infer marginal posterior distributions) or to inform prior distributions. For the present model, I propose to to perform inference on simulated data from a hypothetical scenario. However, real data is used to inform the structure of the prior distribution and the likelihood.

### Zero-sighting surveillance data

As mentioned above, I do not use real data to estimate parameters. Instead, I model a hypothetical situation. The situation is as follows: We assume that at least one fly has been detected; eradication measures have since begun and then ceased; and we now proceed with intensified monitoring, while whatever population that may exist is free to grow relatively unhindered. The goal of the analysis is to infer the probability of eradication for the incipient population, given that no flies detected at any point in this period. Therefore, our "data" is a vector of zeroes, with one for each vector.

Thus, the data we wish to learn from is hypothetical, or simulated. The idea is to simulate a relatively realistic scenario. We observe the outcomes of a surveillance process. The surveillance process is generated by weekly checks of traps that are deployed uniformly in a given area (more about the trapping arrangement below). It is assumed that no specimens are detected at any point in the survey period. In other words, the sum of all detected counts in the period is zero.

### Release recapture studies

In the previous chapter, I briefly discussed release-recapture experiments. These are experiments involving the release of large numbers of flies into networks of standard traps. These studies give us a useful source of information about the probability of capturing a fly given distance between a fly and a trap. 

In cases where Bayesian models have been used, data has not been available on detection rates. For example, @caley2014 and @keith2013 set uninformative priors on the detection rates, and attempt to learn the detection rates for data. However, because of their global economic importance, tephritid fruit flies are relatively well studied. In particular, a number of fruit fly species have been studied with release recapture experiments. Release recapture experiments involve the release of a large number of (often sterilised) fruit flies. These studies help us to learn both the dispersal patterns and tendencies of various species of fly, but also the effectiveness of various trap types and layouts.

## Model

Now that I have discussed background and the data available, I turn to a detailed discussion of the model for this case. This discussion builds on the previous chapter, where the general, basic model was outlined. Here, I focus on the specific prior distributions and likelihood that are used. 

For exposition, I break the model into the following three components: (1) The size of the population (number of individuals); (2) the locations of individuals and traps; and (3) number of individuals caught in traps, conditional on (1) and (2).

The likelihood of a given number of captures is a function of latent variables, namely the number and locations of flies. Under the prior distribution, it is assumed that data at any given timestep are generated as follows. Firstly, nature draws a number of flies (i.e. a population size) which may or may not be based on the number of flies at the previous time step. Next, nature draws a location for each of the flies. Finally, nature draws a number of detections.

### Population size

In this case, $N_1$ is the first week after the most recent fly detection. I have chosen to give $N_1$ the prior distribution $N_1 \mid \lambda \sim \mathrm{Poisson}(\lambda)$, where $\lambda \sim \mathrm{Exponential}(0.05)$. This distribution for $N_1$ is chosen as it is a discrete distribution with right skew, and a relatively large amount of mass $f_{N_1}(x)$ at $x = 0$, corresponding to the situation where flies are already eradicated (see @priorinitsize).

```{r priorinitsize, echo=F, fig.cap = 'Prior distribution of initial population size'}
g = 1/20
init_pmf = function(n, g) g / (1+g)^(n+1)

x = 0:100

df = data.frame(x = x, y = init_pmf(x, g))

ggplot(df) +
  geom_bar(stat = 'identity', aes(x, y)) +
  xlab(bquote(italic(n))) +
  ylab(bquote('Pr('*italic(N)[1] == italic(n)*')'))

# Prior probability that they're already eradicated
```

As for the population sizes at later time points, I assume that growth is exponential at an uncertain rate. In particular, it is assumed that, for $t \in \{2, \ldots, T\}$,  $N_t \mid N_{t-1}, R_t \sim \mathrm{Poisson}(N_{t-1} e^{R_t})$. Here, we can give any continuous distribution for $R_t$. I have chosen the vaguely informative prior $R_t \stackrel{\text{iid}} \sim \mathrm N(0, 0.2)$. The symmetry of this prior means that we are indifferent about whether the population is growing or declining. 

The growth rate prior has been chosen to place the vast majority of density below their estimated growth rate under optimal conditions. Medfly have been estimated to grow at 8% per day, in optimal lab conditions (@papadopoulos2002). Under an exponential growth model this is 56% per week. This can be taken as an upper bound on the growth rate. In the wild, flies may fail to establish due to food scarcity, predation, and/or unsustainably low population density.

In practical, applied cases, it will be desirable to attempt to estimate $R_t$ from data. In particular, it is known that fruit fly population growth rates are highly dependent on temperature. Therefore, if possible, fly population growth rates should be estimated conditional on weather. Choosing an empirically realistic distribution for $R_t$ is likely to improve the efficiency of inference from the survey record. 

### Spatial location

As mentioned in the previous chapter, the model is spatial, insofar as likelihood of detection depends on distances between traps and individual flies.

TODO: Explain that the centre of the grid is the site of the last detection.

#### Population location

The centre of of the population is assumed to be located at the two dimensional vector $L_c$. I have set the prior on $L_c$ to be 
$$
L_c \sim \mathrm {Normal}_2 (\mathbf 0_2, 160^2 I_2),
$$
where $\mathrm{Normal}_2$ is the bivariate normal distribution, $\mathbb 0_2$ is the two-dimensional zero vector and $I_2$ is the $2 \times 2$ identity matrix. This prior reflects a prior belief that the centre of the population is highly likely to be found somewhere on a disc with. E.g., we believe that there is a 95% chance that the centre of the population is within $1.96 \cdot 160 \approx 320$ metres of the centre of the grid.^[For details about how this prior was arrived at, see appendix.] Recall that this makes sense because we believe that the flies are within the distribution.

#### Individual fly dispersals

Let $L_{i, t}$ denote the location of fly $i$ at time $t$, as in the previous chapter. I assign this quantity the prior distribution
$$
L_{i, t} \mid L_c \sim \mathrm {Normal}_2 (L_c, 12.5^2 I_2)
$$
where the notation is defined as in the previous section above. The variance of this distribution is chosen on the basis of prior analysis of dispersal data (see appendix).

Note: Do I need to prove the claims in the next paragraph?

The normal distribution is chosen for a few reasons. Firstly, it is conceptually simple and intuitive to parameterise. Secondly, the location has a simple marginal distribution, thanks to the fact that a normal random variable with a normal mean is itself normal. Thirdly, the distance between a normal random variable and its mean has a known distribution. In particular, for any fly $i$, the squared distance between $L_{i,t}$ and $L_c$, conditional on $L_c$, has the gamma distribution with parameters $\alpha = 1$, and $\beta = 12.5^2 / 2$. Similarly, the distance between $L_{i, t}$ and the origin is the same but with $\beta = (12.5^2 + 160^2) / 2$. Knowing this allows us to easily compare and calibrate the distribution against experimental results. This, in turn, makes elicitation of priors simpler and more intuitive. (It allows us to visualise, for example, the distribution of distances to the origin and to the population centre.)

It may be worth noting that, in real cases, the assumed prior distribution on dispersals may not be reasonable. For example, dispersals may have non-zero mean (due to strength and direction) or non-spherical covariance matrix (i.e. non-equal variances and/or non-zero covariances).^[This is argued by @baker1986.] Further, it may not be reasonable to assume that the fly population cannot move across time. Ideally, the importance of these assumptions should be checked based on the case at hand (e.g. properties of the situation or species in question). Extending the model in various ways to meet these problems is beyond the scope of this work, but warrants further investigation.

### Trap locations

```{r trapgrid, echo=F, fig.cap = 'Illustration of the hypothetical trapping grid. Grey circle represents 200m radius disc surrounding the site of the most most recently detected medfly specimen.'}

grid_size = 4
general_grid_1d = seq(-grid_size / 2, grid_size / 2) * 400
general_grid = expand.grid(general_grid_1d, general_grid_1d)
general_traps = as.matrix(general_grid)

d = 200 / sqrt(2)
supp_grid_1d = seq(-d, d, length.out = 4)
supp_traps = expand.grid(supp_grid_1d, supp_grid_1d)
all_traps = rbind(general_traps, supp_traps)

supp_indicator = rep(c('General', 'Supplementary'), c(nrow(general_traps), nrow(supp_traps)))
traps_df = data.frame(all_traps, factor(supp_indicator, c('Yes', 'No')))

# Plot of traps
ggplot(traps_df) +
  geom_circle(aes(x0=0, y0=0, r=200), colour = 'grey') +
  geom_point(aes(Var1, Var2, col=supp_indicator)) +
  coord_fixed() +
  xlab('X coordinate (metres)') +
  ylab('Y coordinate (metres)') +
  labs(col = 'Trap type') +
  scale_x_continuous(breaks = (-4:4)*200) +
  scale_y_continuous(breaks = (-4:4)*200) 
```

As mentioned previously, it is assumed that trap locations have fixed, known locations. In particular, we assume that monitoring is intensified for the first 6 weeks ($t \in \{1, \ldots, 6\}$). By intensified monitoring, I mean that **supplementary** monitoring traps have been placed alongside the previously existing grid of **general** monitoring traps. More precisely, it is assumed that **general** surveillance traps are placed year round in a 400 $\times$ 400 metre grid (DPIPWE, 2011, p. 50). The **supplementary** surveillance system consists of a set of 16 traps in a circular area, centred at the site of the first fly detection.^[It is typical to wait until at least 2 flies have been detected near each other for an outbreak to be declared. To illustrate the method in a simplified setting, I suppose that one fly detection is sufficient.] See @trapgrid.

#### Probability of capture

I assume that the probability of capture for any given fly in any given trap is 
$$
p(x) = \begin{cases}ax^{-b}, & d>1 \\ a & 0 \geq d \geq 1\end{cases}
$$
where $a = 0.4702111$, $b = 1.37$, and $x$ is the distance between the fly and the trap at the start of the period. Thresholding is introduced because (a) the function does not yield valid probabilities for small enough $x$. In general, the use of a negative exponential function is not ideal here. This function is used for convenience as it is provided by the authors of the study, and data is not publicly accessible. In the context of a more detailed analysis, it may be beneficial to approximate this function with a more "well behaved" function. This is not within the scope of this thesis.

TODO: mention that the probability of capture is adjusted, because the study gives us the lifetime probability of capture.

The results of release recapture studies are highly variable, and gaining access to data may be difficult for older studies. In this case, it may be desirable to incorporate uncertainty about the probability function. This could be done, for example, by assigning prior distributions to the coefficients $a$ and $b$. I do not explore this any further here, as this model exposition is intended for the purposes of illustration only.

## Sampling procedure

In this section, I discuss the problem of computing the posterior distribution, given a survey record. Above, I stated that the model could be defined flexibly. Without restrictions on the form of the growth and detection models, the posterior may be analytically intractable. In other words, we will not be able to write out the posterior density or mass as a function of the data and prior distributions. Such situations are common in the Bayesian framework, because of the tendency for the posterior density or mass to depend, implicitly or explicitly, on analytically intractable integrals.

So far, we have talked about situations when sampling is required for inference. Further problems arise when the model is *agent-based*. In other words, when we include uncertainty about individual-level features in the model. In this case, the detection probability is random, even when the location and population size is known. In other words, the probability of detecting at least one individual is a function of the number of individuals, and also their individual (random) properties. This is a situation in which "the number of things you do not know is one of the things you do not know" (Richardson and Green, 1997).

### Approximate Bayesian Computation

A simple rejection algorithm is used to draw samples from the posterior distribution (see previous chapter). Instead of matching on a statistic, such as the sum of the observations, the rejection rule was set so that data were only kept in cases where the simulated data vector was an exact match with the actual data (the hypothetical zero-catch record).

Here, I have used ABC for intuitiveness, ease of implementation, and the fact that it is relatively efficient for this problem. However, other methods exist that may be worth exploring, for analogous problems where the rejection rate of ABC is higher. There are at least two known methods for sampling from the posterior when the dimension of the parameter space is uncertain. These are the reversible jump MCMC sample (@green1995) and the generalised Gibbs sampler (@keith2015). These may be worth exploring for problems where the rejection rate is high for EBC.

Interestingly, the standard justifications for and against ABC do not apply to the case under consideration. Firstly, the standard justification for ABC is that it allows for inference when the likelihood function is "intractable" - i.e., unknown, uncomputable or otherwise difficult to work with. However, for the current model, the likelihood is known, and relatively simple to write out.

On the other hand, the standard drawback for ABC is that it ensures that we can typically only draw from the posterior approximately. Under standard conditions, we must define a criteria for similarity between simulated and observed data. This is typically done by specifying a summary statistic $S(\mathbb y)$, and a similarity measure $\rho(S(\mathbb y), S(\mathbb y'))$ defined over the space spanning our data $\mathbb y$. We reject a sample if we observe $\rho(\mathbb y_{\text{observed}}, \mathbb y_{\text{simulated}} ) >  \epsilon_0$, where $\epsilon_0$.

$10{,}000$ samples were taken. The acceptance rate for sampling was 0.68. This high level of acceptance is due to the low likelihood of captures, across most of the high prior density region of the model space. The acceptance rate can be interpreted as a numerical approximation to the prior probability of observing zero captures.

## Results

* The fly free period is 12 weeks or 28 days and one generation, whichever is longer (@meats2005). In summer, a Medfly generation takes 28-34 days (@dpirdwa). Therefore, the period I look at is over 12 weeks. However, this may be different based on the period that the manager is interested in.

The posterior probability of extinction after 12 weeks is approximately 0.684.^[Unfortunately, it is not straightforward to visualise the prior or posterior density of this quantity.]

## Discussion

Here I discuss limitations and objections.

### Limitations

* We do not get a posterior distribution over the probability of eradication. 

TODO: Section on objections to/limitations of the model

* Objection: the model is subjective
* Objection: The model is too sensitive to priors. 
  - Defence in Caley 2015
  
### Future work

* Sensitivity analysis for $N_1$
* Sensitivity analysis for growth rates
* Model checking for the growth process
* Estimating growth process parameters from data.
  

TODO: Write conclusion of this chapter

