---
title: "Monthly medfly models"
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Calculating the degree day stopping point for Medfly

The rule is that we stop at $\min(12, g)$ weeks, where $g$ is the number of weeks a "generation" takes, plus 4. the draft medfly CoP uses a degree day formula with average temperatures over 10 years (presumably 1997-2007 or 1998-2008). I have used the table for Adelaide Metro in the back of the draft medfly code of practice. 

I assume we are interested in the probability of eradication for last trap check **before** the CoP says we can reinstate PFA status. The reason for this is that we don't get any new information between then and the PFA reinstatement date (discounting the potential for larval sightings). This means that our informational state is the same on PFA reinstatement day as on the last trap inspection day before PFA reinstatement.


```{r echo=F, include=F}
library(lubridate)
```


```{r}
timelines = read.csv('degree_day_timelines.csv')

knitr::kable(timelines, caption = 'Full table of timelines from the draft medfly CoP')
```

I have manually selected 12 of these to run the simulation for (see below).

# Changes
Changes: Fixed trap locations; larger grid; normally distributed population location.

## Fixed trap locations
```{r}
grid_size = 1
general_grid_1d = seq(-grid_size, grid_size) * 400
general_grid = expand.grid(general_grid_1d, general_grid_1d)
general_traps = as.matrix(general_grid)

par(pty='s')
plot(general_traps)
par(pty='m')

plotrix::draw.circle(0, 0, radius=200, border = 'red' )

d = 200 / sqrt(2)
supp_grid_1d = seq(-d, d, length.out = 4)
supp_grid = expand.grid(supp_grid_1d, supp_grid_1d)
points(supp_grid)
```



```{r}
r = c(0.33, 0.66, 1) * 200
n_traps = c(2, 6, 8)
supp_grid = NULL
for (i in 1:length(r)) {
  rotator = ifelse(i == 1, pi/2, 0)
  
  # Don't bother rotating
  theta = seq(0, 2 * pi, by = 2 * pi / n_traps[i]) 
  new_points = r[i] * cbind(cos(theta), sin(theta))
  supp_grid = rbind(supp_grid, new_points)
}
points(supp_grid)
```

```{r echo=F}
# Set of coordinates for trap locations

# General surveillance
grid_size = 4
general_grid_1d = seq(0, grid_size) * 400
general_grid = expand.grid(general_grid_1d, general_grid_1d)
general_traps = as.matrix(general_grid)

# Supplementary surveillance
grid = expand.grid(seq(-200, 200, 10), seq(-200, 200, 10))
candidates = apply(grid, 1, function(x) if (sum(x^2) < 200^2) x + 400 else NA)
candidates = matrix(na.omit(unlist(candidates)), ncol=2, byrow=T)

# Plot the trap locations
par(pty='s')
plot(c(0, 800), c(0, 800), type = 'n')
points(general_traps, pch=20, col='red', cex=0.1)
points(candidates, pch = 20, cex = 0.1)

N = 1e4
supp_draws = matrix(nrow=N, ncol=4)
n_candidates = nrow(candidates)
for (i in 1:N) {
  supp_draws[i, ] = sort(candidates[sample(1:nrow(candidates), 4, replace = T), 1])
}
means = colMeans(supp_draws)

points(expand.grid(means, means), col = 'red', pch=20)
```

# Simulation code
```{r echo=F}
# Set of coordinates for trap locations
# Specify the indices for the dates we want to check
time_index = c(3, 7, 11, 16, 20, 25, 29, 33, 38, 42, 46, 51)

# Convert to date format
monthly_timelines_list = lapply(timelines[time_index,], as_date)

# Round the end weeks down relative to the start weeks
start = monthly_timelines_list$start
end = monthly_timelines_list$end

# Get number of weeks between start and end, rounded down
difftime_days = ((end - start) - as.numeric(end - start) %% 7)
difftime_weeks = as.difftime(as.numeric(difftime_days) / 7, units='weeks')

MONTHLY_TIMELINES = data.frame(start, end = start + difftime_weeks)

# General surveillance
grid_size = 4
general_grid_1d = seq(-grid_size / 2, grid_size / 2) * 400
general_grid = expand.grid(general_grid_1d, general_grid_1d)
general_traps = as.matrix(general_grid)

d = 200 / sqrt(2)
supp_grid_1d = seq(-d, d, length.out = 4)
supp_traps = expand.grid(supp_grid_1d, supp_grid_1d)
all_traps = rbind(general_traps, supp_traps)

# Plot traps as a check
par(pty = 's')
plot(all_traps)
plotrix::draw.circle(0, 0, radius=200)
par(pty = 'm')

norm = function(x) sqrt(sum(x^2))

pr_fn_ms = function(d) 0.4702111 * d ^(-1.37)
pr_fn_lg = function(d) exp(-2.66076 - 0.01109 * d) / 2

# Get probability of capture as a function of distance
# Inputs
#   d (float): distance btw fly and trap
#   pr_fn (function): one of pr_fn_ms or pr_fn_lg
#   threshold (float): lowest distance for which we will compute the pr_fn function.
pr_cap = function(d, pr_fn, threshold=1) {
  # Probability a single fly IS captured at 1 trap in 1 time period
  # We divide distance by 1000 to rescale to metre scale
  prob = function(distance) 1 - (1 - pr_fn(distance))^(1/4)

  ifelse(d <= threshold, prob(threshold), prob(d))
}

# Function to simulate Medfly population dynamics
# Inputs
#   n_periods (int): number of weeks
#   growth_params (list(mu=numeric(n_periods), sd=numeric(n_periods))): 
#     weekly growth mean and sd
draw_N = function(n_periods, growth_params, type='all') {
  N = growth_rates = numeric(n_periods)
  
  # Draw growth rates
  growth_rates[2:n_periods] = rnorm(n_periods-1, growth_params$mu, growth_params$sd)
  
  # Draw number of flies at t_1
  lambda = rexp(1, 1/20)
  N[1] = rpois(1, lambda) 
  
  # Draw the no. of flies at t_2, t_3, ...
  for (t in 2:n_periods) {
    N[t] = round(N[t-1] * exp(growth_rates[t]))
  }
  
  if (type == 'all') {
    return(N)
  } else {
    return(N[n_periods])
  }
}

# Function to draw a single simulation of a typical scenario
# Inputs
#   n_periods (int): number of weeks
#   growth_params (list(mu=numeric(n_periods), sd=numeric(n_periods))): 
#     weekly growth mean and sd
# Output
#   N (numeric(n_periods)): population size at each (weekly) time point

sim_erad = function(n_periods, growth_params) {
  # Initialise capture data
  y = numeric(n_periods)
  
  # Draw the population size fore each time point
  N = draw_N(n_periods, growth_params)
  
  # Exit early if N[1] == 0 because we already know there will be zero
  # detections.
  if (N[1] == 0) return(0)
  
  # Draw location and individual dispersals/deviations
  L = rnorm(2, sd = 160)
  
  # Capture probabilities
  for (t in 1:n_periods) {
    # Check whether we're in a supplementary or general period
    n_traps = ifelse(t < 7, nrow(all_traps), nrow(supp_traps))
  
    # Draw dispersals
    dispersals = matrix(rnorm(2*N[t], sd=12.5), ncol=2)
    
    # Probability of no captures at t (log scale)
    pr_no_caps_ln = 0
    for (k in 1:n_traps) {
      # Distance between trap k and each fly
      distances = apply(dispersals, 1, function(x) norm(L + x - all_traps[k,]))
      
      # Update probability of no captures
      pr_no_caps_ln = pr_no_caps_ln + sum(log(1-pr_cap(distances, pr_fn_ms)))
    }
    
    #y_t = 1 iff no flies are captured
    y[t] = rbinom(1, 1, exp(pr_no_caps_ln))
    
    if (!y[t]) return(NA)
  }
  
  # Return population sizes and an indicator for acceptance
  # We only accept a sample if y[t] = 1 for all t
  return(tail(N, 1))
}
```


```{r echo=F}
profvis::profvis({
  y = rep(1, n_periods)
  
  # Draw the population size fore each time point
  N = draw_N(n_periods, growth_params)
  
  # Draw location and individual dispersals/deviations
  L = rnorm(2, sd = 160)
  
  # Capture probabilities
  for (t in 1:n_periods) {
    if (N[t] == 0) break
    
    # Check whether we're in a supplementary or general period
    n_traps = ifelse(t < 7, nrow(all_traps), nrow(supp_traps))
  
    # Draw dispersals
    fly_locs = L + matrix(rnorm(2*N[t], sd=12.5), nrow=2)
    
    # Probability of no captures at t (log scale)
    pr_no_caps_ln = 0
    for (k in 1:n_traps) {
      # Distance between trap k and each fly
      distances = sqrt(colSums((fly_locs - all_traps[k,])^2))
      
      # Update probability of no captures
      pr_no_caps_ln = pr_no_caps_ln + sum(log(1-pr_cap(distances, pr_fn_ms)))
    }
    
    #y_t = 1 iff no flies are captured
    y[t] = rbinom(1, 1, exp(pr_no_caps_ln))
  
})
```


```{r echo=F}
```


```{r echo = F}

# Draw 1e4 posterior samples
sim_post = function(n_iter, n_periods, growth_params) {
  posterior_N = numeric(n_iter)
  count_success = count_all = 1
  
  pb = progress::progress_bar$new(total=n_iter)
  
  while (count_success < n_iter) {
    new = sim_erad(n_periods, growth_params)
    
    if (!is.na(new)) {
      posterior_N[count_success] = new
      pb$tick()
      count_success = count_success + 1
    }
    
    count_all = count_all + 1
  }
  
  print(paste('Success rate:', count_success / count_all))
  
  return(posterior_N)
}

# Draw from the prior
sim_prior = function(n_iter, n_periods, growth_params) {
  # Init N
  prior_N = numeric(n_iter)
  
  for (i in 1:n_iter) {
    prior_N[i] = draw_N(n_periods, growth_params, 'last')
  }
  
  return(prior_N)
}
```

```{r echo=F}
MU_R = c(-0.067, -0.015, 0.028, -0.001, -0.009, 0.011, 
          -0.014, -0.008, 0.008, 0.015, 0.009, -0.005)

SD_R = sqrt(c(0.028, 0.022, 0.028, 0.02, 0.02, 0.018, 
            0.016, 0.016, 0.012, 0.013, 0.033, 0.019))

sim_all_months = function(n_iter) {
  post_prob = numeric(12)
  prior_prob = numeric(12)
  
  for (i in 1:nrow(MONTHLY_TIMELINES)) {
    dd_weeks = diff(unlist(MONTHLY_TIMELINES[i,])) / 7
    n_periods = max(dd_weeks, 12)
    
    start_date = MONTHLY_TIMELINES[i, 1]
    end_date = start_date + 7 * n_periods - 1
    
    print(paste('Start date:', start_date))
    print(paste('End date:  ', end_date))
    
    # Get prior growth parameters
    # Get the month of each day in the observation period
    months = month(seq(start_date, end_date, 1))
    
    
    # Get the parameters corresponding to the months
    mu = matrix(MU_R[months], nrow=7)
    sd = matrix(SD_R[months], nrow=7)

    growth_params = list(mu_r = colSums(mu),
                         sd_r = colSums(sd))
    
    # Draw priors
    prior_draws = sim_prior(n_iter, n_periods, growth_params)
    prior_prob[i] = mean(prior_draws == 0)
    
    
    # Draw posteriors
    post_draws = sim_post(n_iter, n_periods, growth_params)
    post_prob[i] = mean(post_draws == 0)
  }
  
  return(list(post=post_prob, prior=prior_prob))
}
```

```{r echo = F}
if (FALSE) {
  results = sim_all_months(1e4)
  saveRDS(results, 'monthly_results.RDS') 
}

results = readRDS('monthly_results.RDS')
# print(results)
```

```{r echo = F}
output = data.frame(MONTHLY_TIMELINES, results)
colnames(output) = c("Start date",
                     "Final inspection before PFA reinstatement",
                     "Posterior prob. of erad.",
                     "Prior prob. of erad.")

knitr::kable(output, caption = 'Table of results for all simulations')
```

